---
pagetitle: "AGUASAN 2027 topic pitch - Publishing WASH Data Together (person story)"
format:
  revealjs:
    theme: [default, custom.scss]
    width: 1250
    height: 720
    margin: 0.07
    transition: none
    slide-number: true
    progress: false
    controls: false
    center: false
    auto-stretch: false
    hash: true
    filters:
      - roughnotation
    include-after-body:
      text: |
        <script>
        // The roughnotation extension only draws .rn annotations on the R
        // key; draw them automatically when their slide is shown instead
        // (the a/b/c kicker boxes rely on this).
        window.addEventListener("load", function () {
          if (!window.Reveal || !window.RoughNotation) return;
          var annotated = new Map();
          function annotateCurrent(slide) {
            if (!slide) return;
            slide.querySelectorAll(".rn").forEach(function (el) {
              if (!annotated.has(el)) {
                annotated.set(el, RoughNotation.annotate(el, {
                  type: el.dataset.rnType || "box",
                  animate: true,
                  animationDuration: 600,
                  color: el.dataset.rnColor || "#5b195b",
                  strokeWidth: parseInt(el.dataset.rnStrokewidth, 10) || 2,
                  padding: parseInt(el.dataset.rnPadding, 10) || 5,
                  multiline: true,
                  iterations: 2
                }));
              }
              var a = annotated.get(el);
              a.show();
              if (a._svg) {
                a._svg.style.transform = "scale(" + 1 / Reveal.getScale() + ")";
                a._svg.style.transformOrigin = "top left";
              }
            });
          }
          Reveal.on("slidechanged", function (e) { annotateCurrent(e.currentSlide); });
          Reveal.on("ready", function (e) { annotateCurrent(e.currentSlide); });
          if (Reveal.isReady && Reveal.isReady()) {
            annotateCurrent(Reveal.getCurrentSlide());
          }
        });
        </script>
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(stringr)
  library(ggplot2)
  library(forcats)
})

# Local clone of openwashdata/washopenresearch (read-only here).
pkg_path <- "/Users/lschoebitz/Documents/gitrepos/gh-org-openwashdata/washopenresearch"
suppressMessages(devtools::load_all(pkg_path, quiet = TRUE))

ink <- "#1e1e1e"
paper <- "#f5eef5"
accent <- "#5b195b"
muted <- "#6b665e"

fmt <- function(x) format(x, big.mark = ",", trim = TRUE)
pct1 <- function(n, d) {
  paste0(format(round(100 * n / d, 1), nsmall = 1, trim = TRUE), "%")
}
pct0 <- function(n, d) paste0(round(100 * n / d), "%")

# Two journals ship as exported package datasets (washdev, ploswater); the
# other two as raw IWA snapshots in the package's data-raw/ folder.
read_snap <- function(file) {
  suppressWarnings(read_csv(
    file.path(pkg_path, "data-raw", file),
    show_col_types = FALSE,
    name_repair = "unique_quiet"
  ))
}

ws <- read_snap("ws.csv")
jwh <- read_snap("jwh.csv")

n_screened <- nrow(ws) + nrow(jwh) + nrow(washdev) + nrow(ploswater)

# A statement counts as "reaches an open repository" when it names one.
# washdev and ploswater carry a normalised das_type; ws and jwh carry the
# raw statement text, so one pattern covers both forms.
open_pattern <- regex(
  "repositor|genbank|dataverse|ncbi|zenodo|figshare|dryad|dhsprogram|publicly available in",
  ignore_case = TRUE
)

prep <- function(d) {
  d |>
    transmute(
      das_type = as.character(das_type),
      published_year = as.integer(published_year)
    )
}

journal_levels <- c(
  "Water Supply",
  "J. Water and Health",
  "J. WASH for Development",
  "PLOS Water (research articles)"
)

das_levels <- c(
  "Open repository", "In paper or supplement", "On request",
  "No statement or other"
)

papers <- bind_rows(
  "Water Supply" = prep(ws),
  "J. Water and Health" = prep(jwh),
  "J. WASH for Development" = prep(washdev),
  "PLOS Water (research articles)" =
    prep(ploswater |> filter(article_type == "Research Article")),
  .id = "journal"
) |>
  mutate(
    journal = factor(journal, levels = journal_levels),
    open = !is.na(das_type) & str_detect(das_type, open_pattern),
    das4 = case_when(
      open ~ "Open repository",
      !is.na(das_type) & str_detect(das_type, regex(
        "request|contact the corresponding|not shareable",
        ignore_case = TRUE
      )) ~ "On request",
      !is.na(das_type) & str_detect(das_type, regex(
        "^in paper$|included in the (paper|article)|provided in th|supplementary",
        ignore_case = TRUE
      )) ~ "In paper or supplement",
      TRUE ~ "No statement or other"
    ),
    das4 = factor(das4, levels = das_levels)
  )

n_plotted <- nrow(papers)

by_journal <- papers |>
  group_by(journal) |>
  summarise(total = n(), n_open = sum(open), .groups = "drop") |>
  mutate(share = n_open / total)

yr_all_min <- min(papers$published_year, na.rm = TRUE)
yr_all_max <- max(papers$published_year, na.rm = TRUE)

# The three journals without an enforced data policy vs the one with it.
enforced <- "PLOS Water (research articles)"
noenf <- by_journal |> filter(journal != enforced)
n_noenf <- sum(noenf$total)
n_noenf_open <- sum(noenf$n_open)
one_in_noenf <- round(n_noenf / n_noenf_open)

plos <- by_journal |> filter(journal == enforced)
one_in_plos <- round(plos$total / plos$n_open)

# All four DAS categories per journal, 100 percent stacked; the open
# repository segment leads the stack and carries the accent colour.
das_by_journal <- papers |>
  count(journal, das4, .drop = FALSE) |>
  group_by(journal) |>
  mutate(total = sum(n), share = n / total) |>
  ungroup() |>
  left_join(by_journal |> select(journal, open_share = share), by = "journal") |>
  mutate(journal_wrapped = str_replace(journal, " \\(", "\n("))

das_fill <- c(
  "Open repository" = accent,
  "In paper or supplement" = "#8a8378",
  "On request" = "#b3ada3",
  "No statement or other" = "#dedad3"
)

alt_routes <- paste0(
  "Stacked horizontal bar chart: for each of four WASH journals, the share ",
  "of articles in each data availability statement category (open ",
  "repository, in paper or supplement, on request, no statement or other). ",
  "Open repository shares: ",
  paste0(
    by_journal$journal, " ", pct0(by_journal$n_open, by_journal$total),
    collapse = ", "
  ),
  ". The three journals without an enforced data policy stay below 4 ",
  "percent; PLOS Water, which requires and supports the statement, is ",
  "above 50 percent."
)

plot_routes <- function(base_size, label_size, show_caption) {
  df <- das_by_journal |>
    mutate(
      journal_wrapped = fct_reorder(journal_wrapped, open_share),
      seg_label = pct0(n, total),
      light_fill = das4 %in% c("On request", "No statement or other")
    ) |>
    # Label midpoints from the full stack, so filtering small segments out
    # of the label layer cannot shift the remaining labels.
    arrange(journal, das4) |>
    group_by(journal) |>
    mutate(x_mid = cumsum(share) - share / 2) |>
    ungroup()
  p <- df |>
    ggplot(aes(x = share, y = journal_wrapped, fill = das4)) +
    geom_col(width = 0.68, position = position_stack(reverse = TRUE)) +
    geom_text(
      data = df |> filter(share >= 0.05),
      aes(x = x_mid, label = seg_label, colour = light_fill),
      size = label_size, fontface = "bold"
    ) +
    geom_label(
      data = df |> filter(das4 == "Open repository", share < 0.05),
      aes(x = share + 0.004, label = seg_label),
      hjust = 0, colour = accent, fill = "#ffffff", alpha = 0.92,
      size = label_size, fontface = "bold",
      label.size = 0, label.padding = unit(0.14, "lines")
    ) +
    scale_fill_manual(values = das_fill, name = NULL) +
    scale_colour_manual(
      values = c(`TRUE` = ink, `FALSE` = paper), guide = "none"
    ) +
    scale_x_continuous(limits = c(0, 1.001), expand = c(0, 0)) +
    labs(
      x = NULL, y = NULL,
      title = paste0(
        "Data almost never reaches an open repository, ",
        "unless the journal asks and supports"
      ),
      subtitle = paste0(
        "Data availability statements by journal, share of ",
        fmt(n_plotted), " articles, ", yr_all_min, " to ", yr_all_max
      )
    ) +
    guides(fill = guide_legend(nrow = 1)) +
    theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      plot.background = element_rect(fill = paper, colour = NA),
      panel.background = element_rect(fill = paper, colour = NA),
      panel.grid = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(colour = ink, lineheight = 0.92),
      plot.title = element_text(face = "bold", size = base_size * 0.78),
      plot.subtitle = element_text(colour = muted, size = base_size * 0.64),
      plot.title.position = "plot",
      legend.position = "top",
      legend.location = "plot",
      legend.justification = "left",
      legend.margin = margin(2, 0, 2, 0),
      legend.text = element_text(size = base_size * 0.62),
      legend.key.size = unit(0.55, "cm"),
      plot.caption = element_text(colour = muted, size = 12, hjust = 0),
      plot.caption.position = "plot",
      plot.margin = margin(4, 10, 2, 4)
    )
  if (show_caption) {
    p <- p + labs(caption = paste0(
      "Data: https://github.com/openwashdata/washopenresearch (CC BY 4.0)."
    ))
  }
  p
}

# Supplementary file formats behind the "available" statements. Multi-label:
# one paper can attach several formats, so shares are of papers WITH
# supplements and do not sum to 100. Tokens follow the owner's draft-deck
# approach: split on separators, exact-match tokens plus mime fragments
# (ploswater carries free-text junk, so bare substring matching overcounts).
supp_cols <- function(d) {
  d |>
    transmute(
      is_supp = as.logical(is_supp),
      supp_file_type = as.character(supp_file_type)
    )
}

supp <- bind_rows(
  supp_cols(washdev), supp_cols(ploswater), supp_cols(uncnewsletter)
) |>
  filter(is_supp %in% TRUE) |>
  mutate(tokens = purrr::map(
    supp_file_type,
    \(x) str_squish(str_to_lower(unlist(str_split(x, "[;,]"))))
  ))

n_supp <- nrow(supp)
has_tok <- function(pred) {
  sum(purrr::map_lgl(supp$tokens, \(t) any(pred(t), na.rm = TRUE)))
}
n_fmt_word <- has_tok(\(t) t %in% c("doc", "docx") |
  str_detect(t, "wordprocessingml|msword"))
n_fmt_pdf <- has_tok(\(t) str_detect(t, "\\bpdf\\b"))
n_fmt_excel <- has_tok(\(t) t %in% c("xls", "xlsx") |
  str_detect(t, "spreadsheetml|ms-excel"))
n_fmt_img <- has_tok(\(t) t %in% c("png", "jpg", "jpeg", "tif", "tiff", "gif") |
  str_detect(t, "image/"))
n_fmt_csv <- has_tok(\(t) t == "csv" | str_detect(t, "\\bcsv\\b"))

formats_df <- tibble::tibble(
  format = c("Word document", "PDF", "Excel", "Image", "CSV"),
  n = c(n_fmt_word, n_fmt_pdf, n_fmt_excel, n_fmt_img, n_fmt_csv)
) |>
  mutate(
    share = n / n_supp,
    qualifies = format == "CSV",
    bar_label = pct0(n, n_supp)
  )

alt_formats <- paste0(
  "Horizontal bar chart of the file formats attached as supplementary ",
  "data by the ", fmt(n_supp), " papers with supplements: ",
  paste0(
    formats_df$format, " ", pct1(formats_df$n, n_supp),
    collapse = ", "
  ),
  ". Only the CSV bar, the one machine-readable non-proprietary format, ",
  "is drawn in the accent colour; a paper can attach several formats."
)

plot_formats <- function(base_size, label_size) {
  formats_df |>
    ggplot(aes(
      x = share,
      y = fct_reorder(format, share),
      fill = qualifies
    )) +
    geom_col(width = 0.68) +
    geom_text(
      aes(label = bar_label),
      hjust = -0.15, colour = ink, size = label_size, fontface = "bold"
    ) +
    scale_fill_manual(
      values = c(`TRUE` = accent, `FALSE` = muted), guide = "none"
    ) +
    scale_x_continuous(
      limits = c(0, max(formats_df$share) * 1.18), expand = c(0, 0)
    ) +
    labs(
      x = NULL, y = NULL,
      title = "The attached \"data\" is mostly Word and PDF files",
      subtitle = paste0(
        "Share of the ", fmt(n_supp), " papers with supplementary files; ",
        "one paper can attach several formats"
      ),
      caption = paste0(
        "Data: https://github.com/openwashdata/washopenresearch (CC BY 4.0)."
      )
    ) +
    theme_minimal(base_size = base_size) +
    theme(
      text = element_text(colour = ink),
      plot.background = element_rect(fill = paper, colour = NA),
      panel.background = element_rect(fill = paper, colour = NA),
      panel.grid = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(colour = ink),
      plot.title = element_text(face = "bold", size = base_size * 0.78),
      plot.subtitle = element_text(colour = muted, size = base_size * 0.64),
      plot.title.position = "plot",
      plot.caption = element_text(colour = muted, size = 12, hjust = 0),
      plot.caption.position = "plot",
      plot.margin = margin(4, 10, 2, 4)
    )
}

```

::: {.kicker}
AGUASAN Workshop 2027, topic pitch
:::

::: {.display}
Meet Alex.
:::

Fifteen years in WASH programmes. The funding just collapsed. Years of collected data sit on her desktop.

::: {.muted}
Alex is a composite, drawn from the practitioners in our two course cohorts. She is not one real person. Her situation is real everywhere.
:::

::: {.topic}
Our topic: [Publishing WASH Data Together]{.accent}, a hands-on retreat for an open, AI-ready sector.
:::

::: {.foot}
Elizabeth Tilley and Lars Schöbitz, Global Health Engineering, ETH Zurich
:::

::: {.notes}
Controlling idea: the sector can no longer afford closed data, and one retreat week turns a practitioner hit by the funding collapse into a published, citable data author with skills that outlast any WASH job.

Budget 0:35, cumulative 0:35.

Thank you, Roger.

Everyone in this room is managing the same issues: budgets cut, programmes closing, good people leaving the sector.

So let me pitch our topic through one of those people. I call her Alex and she's a composite of the practitioners in our two course cohorts.

Fifteen years in WASH programmes. Last year the funding collapsed and years of collected data sit on her desktop: surveys, monitoring rounds, a thesis.

Our proposal, Publishing WASH Data Together, is about what happens next, for Alex and for that data.
:::

## Where is Alex's data supposed to go? {#why-now}

::: {.kicker-q}
[a\) Why this topic, why now]{.rn rn-type="box" rn-color="#5b195b" rn-strokewidth="2" rn-padding="6"}
:::

```{r}
#| label: chart-routes
#| fig-width: 13
#| fig-height: 4.4
#| fig-dpi: 160
#| out-width: "100%"
#| fig-alt: !expr alt_routes

plot_routes(base_size = 22, label_size = 6.6, show_caption = TRUE)
```

Without a journal that asks and supports: [1 in `r one_in_noenf`]{.rn-fragment rn-type="circle" rn-color="#5b195b" rn-strokewidth="3" rn-padding="10"} articles points to data in an open repository.

::: {.notes}
Budget 0:45, cumulative 1:20.

Question a: why now? Let's start with where Alex's data could even go. We screened about `r fmt(n_screened)` articles across four WASH journals.

Best practice is an open repository, separate from the article. PLOS Water has an enforced data policy, resulting in 1 of 2 papers deposited. Our proposal is exactly that: ask and support.

In the three journals without an enforced data policy: CLICK. One in `r one_in_noenf` articles points to an open repository. SLOW DOWN on this number, it carries the circle, let it land. BREATH POINT.

And journals are only what is visible. Most data, like Alex's, never reaches a paper at all. It's no unwillingness: no credit, no time, no reward.

:::

## When files are attached, what are they? {#formats}

::: {.kicker-q}
[a\) Why this topic, why now]{.rn rn-type="box" rn-color="#5b195b" rn-strokewidth="2" rn-padding="6"}
:::

```{r}
#| label: chart-formats
#| fig-width: 13
#| fig-height: 4.8
#| fig-dpi: 160
#| out-width: "100%"
#| fig-alt: !expr alt_formats

plot_formats(base_size = 22, label_size = 6.6)
```

Only CSV is machine readable and non proprietary. [A CSV without documentation is still closed data]{.rn-fragment rn-type="underline" rn-color="#5b195b" rn-strokewidth="3"}.

::: {.notes}
Budget 0:30, cumulative 1:50.

When articles do attach "data", it is mostly Word and PDF: tables your eyes can read and no computer can. The one format that qualifies, machine readable and non proprietary, is CSV: 2 percent.

CLICK for the underline.

And a CSV without documentation is still closed data. Documented, described, reusable: that is what the retreat teaches.

OPTIONAL BEAT, if pace allows, otherwise hold for the Q&A window:

And before anyone credits the repository pointers from the last slide, we followed every one of them; about one in five is dead, and most deposits we could open hold archives with no machine-readable file and no README. The format and documentation problem follows the data wherever it goes.
:::

## Why AGUASAN, and why 2027 {#why-aguasan}

::: {.kicker-q}
[a\) Why as an AGUASAN Workshop]{.rn rn-type="box" rn-color="#5b195b" rn-strokewidth="2" rn-padding="6"}
:::

- **The 40th AGUASAN Workshop:** the community of practice, hosted by the SDC, that shapes the WASH sector's practice, and the participants are the people who hold the data
- **The retreat format is the method:** a week of hands-on practical workshops, not a lecture series
- **The funding window closes in 2026:** five ETH Board Open Research Data projects, about CHF 500,000, the 2024 National Prize, 51 published datasets; AGUASAN is the bridge from a funded programme to standing sector practice

::: {.notes}
Budget 0:30, cumulative 2:20.

Still question a: why AGUASAN, why 2027? The timing will not repeat. Five ETH Board Open Research Data projects with about half a million francs in funding. We have built the machinery: a community, workflows, tools, a free academy, and 51 published datasets. 

That funding window closes in 2026. AGUASAN buys us the room and the reach.
:::

## Alex signs up: the work starts four weeks early {#prework}

::: {.kicker-q}
[b\) What participants take away]{.rn rn-type="box" rn-color="#272bd1" rn-strokewidth="2" rn-padding="6"}
:::

- **Months before the workshop:** the signup form asks for her dataset, screened before acceptance; data is the entry ticket
- **At least four weeks of pre-work with individual support:** accounts, setup, documentation, permissions
- **No coding background required:** designed for novices, and a shared cloud environment gives everyone the same start
- **Qualitative data is welcome:** the FAIRqual project built the experience for the harder cases

::: {.notes}
Budget 0:25, cumulative 2:45.

Now question b: what does Alex take away? Months before the workshop, the signup form asks for her dataset, which is screened before acceptance. The data is the entry ticket and any dataset counts: surveys, monitoring rounds, lab results, and qualitative material, such as audio recordings.

Then at least four weeks of supported pre-work: accounts, setup, documentation, permissions, all settled before arrival. Nobody arrives as a coder; the week is designed for novices.
:::

## What Alex takes home {#takeaway}

::: {.kicker-q}
[b\) What participants take away]{.rn rn-type="box" rn-color="#272bd1" rn-strokewidth="2" rn-padding="6"}
:::

- **A published data package:** documented, openly licensed, citable with a DOI; final draft by Friday, archiving the final step
- **[Skills that outlast any WASH job]{.rn-fragment rn-type="underline" rn-color="#5b195b" rn-strokewidth="3"}:** Git and GitHub, Quarto (alternative for MS Word), safe and confident work with AI agents
- **A plan:** each participant leaves with concrete next steps for their own organisation

::: {.notes}
Budget 0:35, cumulative 3:20.

Then Friday. What flies home with Alex? First, a data package: documented, openly licensed, citable, in final state by Friday. Her data stops being a folder on a desktop. BREATH POINT.

Second, the heart of it: skills that outlast any WASH job. CLICK. Git and GitHub, Quarto, which is an alternative for MS Word, and safe, confident work with AI agents. When funding shrinks, these skills keep people employable, inside the sector or beyond it.
:::

## We carry it: before, during, after {#custodian}

::: {.kicker-q}
[c\) Who convenes and follows up]{.rn rn-type="box" rn-color="#c14a09" rn-strokewidth="2" rn-padding="6"}
:::

- **Custodians:** Elizabeth Tilley and Lars Schöbitz, Global Health Engineering, ETH Zurich
- **Certified teaching:** Adriana Clavijo, full-time Scientific Assistant, Carpentries instructor; Lars, RStudio instructor; evidence-based, novice-first
- **Committed to all of it:** October prep event, Steering Committee, case studies, presenting
- **A permanent ETH position** carries the community beyond any grant
- **Follow-up:** [a free third data science course teaching foundational R skills]{.rn-fragment rn-type="underline" rn-color="#5b195b" rn-strokewidth="3"}

::: {.notes}
Budget 0:35, cumulative 3:55.

The last question: who carries it? That's us, Liz and I, from Global Health Engineering, as custodians, committed to every contribution in the template, from the October preparatory event to the Steering Committee.

Our teaching is professional: Adriana Clavijo is a full time Scientific Assistant in our open science team. We are both certified instructors, and teach with evidence-based methods built for novices.

My permanent ETH position carries the community beyond any single grant and we will offer a concrete follow-up for all participants. CLICK. A free third data science course teaching foundational R skills, where participants mix with others from across the sector.
:::

## Thanks [{{< fa sun >}}]{.accent} {#thanks}

::: {.columns}
::: {.column width="47%"}

::: {.kicker}
References
:::

- **Data and code:** [github.com/openwashdata/washopenresearch](https://github.com/openwashdata/washopenresearch)
- **These slides:** [download as PDF](https://github.com/Global-Health-Engineering/aguasan/raw/main/slides/aguasan-2027-topic-pitch-openwashdata-lschoebitz.pdf)
- **Built with Quarto and reveal.js:** [quarto.org/docs/presentations/revealjs](https://quarto.org/docs/presentations/revealjs/)

:::
::: {.column width="53%"}

::: {.kicker}
Resources
:::

- **The openwashdata community:** [openwashdata.org](https://openwashdata.org)
- **The FAIRqual project:** [fairqual.org](https://fairqual.org)
- **Two published dataset examples:** [fecalcanuga](https://openwashdata.github.io/fecalcanuga/) & [solidwastekampala](https://openwashdata.github.io/solidwastekampala/) and see [catalogue for all](https://openwashdata.org/pages/gallery/data/)
- **Our newsletter (250 subscribers):** [buttondown.email/openwashdata](https://buttondown.email/openwashdata)

:::
:::

::: {.foot}
All material is licensed under Creative Commons Attribution 4.0 International (CC BY 4.0).
:::

::: {.notes}
Budget 0:30, cumulative 4:25.

Thank you. Everything behind this pitch is open. Let's send fifty Alexes home with published data."
:::
